---
title: "Time Series Analysis and Forecasting of Air Passengers"
author: Donnie
runtime: shiny
format: 
  html:
    code-tools: true
    code-fold: false
    execute:
      echo: true
      warning: false
      message: false
      cache: false
---


## Abstract
This report presents a comprehensive and mathematically detailed time series analysis...


## 1. Introduction

Forecasting monthly airline passenger volumes is essential for planning, scheduling, and resource allocation. 
This report demonstrates a complete, mathematically rigorous workflow following the SORS6102 project brief.

The objectives include:

1. Data description and preparation  
2. Exploratory analysis (trend, seasonality, variance)  
3. Stationarity testing  
4. Model identification (ACF/PACF, differencing)  
5. ARIMA/SARIMA estimation  
6. Residual diagnostics  
7. Forecasting (with back-transformation)  
8. Evaluation of accuracy  
9. Interpretation and recommendations  

---

## 2. Data and Setup

```{r}
library(forecast)
library(tseries)
library(ggplot2)
library(lmtest)
library(kableExtra)
library(dplyr)

data("AirPassengers")
ts_data <- AirPassengers
```
```{r libraries, message=FALSE, warning=FALSE}
# Required libraries
library(ggplot2)
library(viridis)       # colorblind-friendly palette
library(extrafont)     # for Times New Roman fonts

# If running for the first time (uncomment)
# extrafont::font_import()
# extrafont::loadfonts(device="pdf")

# Load data
data("AirPassengers")
ts_data <- AirPassengers
```
The dataset consists of **144 monthly observations** from **Jan 1949 – Dec 1960**, with frequency 12.

A raw plot shows increasing variance and multiplicative seasonality:

```{r}
autoplot(ts_data) +
  ggtitle("Monthly AirPassengers") +
  ylab("Passengers")
```



```{r plot, message=FALSE, warning=FALSE}
library(ggplot2)
library(viridis)

data("AirPassengers")
ts_data <- AirPassengers

ggplot(
    data = NULL,
    aes(x = as.numeric(time(ts_data)),
        y = as.numeric(ts_data))
  ) +
  geom_line(color = viridis(1, option = "C"), size = 1.1) +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    axis.title = element_text(face = "bold", size = 14),
    axis.text  = element_text(size = 12),
    axis.line  = element_line(color = "black", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85"),
    legend.position = "none",
    plot.margin = margin(10,10,10,10)
  ) +
  labs(
    x = "Year",
    y = "Number of Passengers"
  )
```



---

## 3. Transformation

Variance increases with level, so we apply a log transform:

$$
Y_t = \log(X_t)
$$

```{r}
log_ts <- log(ts_data)
autoplot(log_ts)
```
Or 

```{r logplot, message=FALSE, warning=FALSE}
log_ts <- log(ts_data)

autoplot(log_ts) +
  geom_line(color = viridis(1, option = "C"), linewidth = 1.1) +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    axis.title = element_text(face = "bold", size = 14),
    axis.text  = element_text(size = 12),
    axis.line  = element_line(color = "black", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85"),
    legend.position = "none",
    plot.margin = margin(10,10,10,10)
  ) +
  labs(
    x = "Year",
    y = "Log(Passengers)"
  )
```
---

## 4. STL Decomposition

```{r}
stl_fit <- stl(log_ts, s.window="periodic")
autoplot(stl_fit)
```

Or 

```{r stlplot, message=FALSE, warning=FALSE}

stl_fit <- stl(log_ts, s.window = "periodic")

autoplot(stl_fit) +
  scale_color_viridis_d(option = "C") +
  scale_fill_viridis_d(option = "C") +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    strip.background = element_rect(fill = "white"),
    strip.text       = element_text(face = "bold", size = 13, family = "serif"),
    axis.title       = element_text(face = "bold", size = 14),
    axis.text        = element_text(size = 12),
    axis.line        = element_line(color = "black", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85"),
    legend.position  = "none",
    plot.margin      = margin(10,10,10,10)
  ) +
  labs(
    x = "Year",
    y = "Value"
  )
```

Trend and seasonal components are clearly visible.

---

## 5. Stationarity and Differencing

ADF test on the logged data:

```{r}
adf.test(log_ts)
```

Not stationary → apply first and seasonal differencing:

$$
\nabla (1 - B^{12}) Y_t = (1 - B)(1 - B^{12}) Y_t
$$

```{r}
d1s <- diff(diff(log_ts), lag=12)
adf.test(d1s)
```

---


## 6. ACF and PACF Analysis

To identify ARIMA orders, we examine ACF and PACF of the differenced series:

```{r}
par(mfrow=c(1,2))
Acf(d1s, main="ACF of Differenced Series")
Pacf(d1s, main="PACF of Differenced Series")
par(mfrow=c(1,1))
```

Or 

```{r acf_pacf, message=FALSE, warning=FALSE}

# ACF plot
p_acf <- ggAcf(d1s) +
  scale_color_viridis_d(option = "C") +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    axis.title = element_text(face = "bold", size = 14),
    axis.text  = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85")
  ) +
  labs(title = "ACF of Differenced Series")

# PACF plot
p_pacf <- ggPacf(d1s) +
  scale_color_viridis_d(option = "C") +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    axis.title = element_text(face = "bold", size = 14),
    axis.text  = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85")
  ) +
  labs(title = "PACF of Differenced Series")

# Display side-by-side
library(gridExtra)
grid.arrange(p_acf, p_pacf, ncol = 2)
```

**Interpretation:**

- Seasonal spikes at lag 12 suggest seasonal differencing was needed.
- ACF tailing + PACF significant at lag 1 suggests an MA(1) component.
- Seasonal MA(1) at lag 12 is also likely.

These patterns support the *Airline Model*:

$$
\text{ARIMA}(0,1,1)(0,1,1)_{12}
$$

---

## 7. Model Estimation

### 7.1 Fit the canonical Airline Model

```{r}
fit_airline <- Arima(log_ts, order=c(0,1,1), seasonal=c(0,1,1))
summary(fit_airline)
```

Model equations:

$$
(1 - B)(1 - B^{12})Y_t = (1 + \theta_1 B)(1 + \Theta_1 B^{12}) \varepsilon_t
$$

Where:

- $\theta_1$ = non-seasonal MA(1)
- $\Theta_1$ = seasonal MA(1)

---

## 7.2 Fit alternative SARIMA models

```{r}
fit_alt1 <- Arima(log_ts, order=c(2,1,1), seasonal=c(0,1,1))
fit_alt2 <- Arima(log_ts, order=c(0,1,1), seasonal=c(1,1,1))
```

---

## 7.3 Model Comparison

```{r}
models <- list(
  Airline = fit_airline,
  Alt1 = fit_alt1,
  Alt2 = fit_alt2
)

model_table <- data.frame(
  Model = names(models),
  AIC  = sapply(models, AIC),
  # AICc = sapply(models, forecast::AICc),
  BIC  = sapply(models, BIC)
)

kable(model_table, caption="Model Selection Criteria") %>% kable_styling(full_width=FALSE)
```
```{r}
library(forecast)

fit_alt1 <- Arima(log_ts, order=c(2,1,1), seasonal=c(0,1,1))
fit_alt2 <- Arima(log_ts, order=c(0,1,1), seasonal=c(1,1,1))

models <- list(
  Airline = fit_airline,
  Alt1 = fit_alt1,
  Alt2 = fit_alt2
)

model_table <- data.frame(
  Model = names(models),
  AIC  = sapply(models, AIC),
  # AICc = sapply(models, forecast::AICc),
  BIC  = sapply(models, BIC)
)

model_table
```

**Conclusion:**  
The Airline Model often achieves the lowest AICc and is preferred for this dataset.

---

## 8. Residual Diagnostics

```{r residuals, message=FALSE, warning=FALSE}
# Residuals
res <- residuals(fit_airline)

# --- 1. Residual Time Series Plot ---
p1 <- autoplot(res) +
  geom_line(color = viridis(1, option = "C"), linewidth = 1) +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85")
  ) +
  labs(x = "Time", y = "Residuals")

# --- 2. ACF Plot ---
p2 <- ggAcf(res) +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85")
  ) +
  labs(title = "ACF of Residuals")

# --- 3. Histogram with Density ---
p3 <- ggplot(data.frame(res), aes(x = res)) +
  geom_histogram(aes(y = ..density..),
                 bins = 20,
                 fill = viridis(1, option="C"), 
                 color = "white",
                 alpha = 0.9) +
  geom_density(color = "black", linewidth = 1) +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    axis.title = element_text(face = "bold", size = 14),
    axis.text  = element_text(size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray85")
  ) +
  labs(x = "Residuals", y = "Density")

# Arrange all three
library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 3)

# Print Ljung-Box Test:
Box.test(res, lag = 24, type = "Ljung-Box", fitdf = 2)
```

### Requirements for a valid model:

1. Residuals ≈ white noise  
2. Ljung–Box p-value > 0.05  
3. No seasonal structure in residuals  
4. Approximate normality  

```{r}
Box.test(residuals(fit_airline), lag=24, type="Ljung-Box", fitdf=2)
```

If p-value > 0.05, no remaining autocorrelation.

---

## 9. Forecasting

We forecast in **log scale**, then back-transform.

```{r forecast_plot, message=FALSE, warning=FALSE}


# Forecast (already in log scale)
fcast <- forecast(fit_airline, h = 24, level = c(80, 95))

# Convert autoplot output into ggplot and restyle it
autoplot(fcast) +
  scale_color_viridis_d(option = "C") +
  scale_fill_viridis_d(option = "C", alpha = 0.6) +
  theme_minimal(base_family = "serif") +
  theme(
    panel.background  = element_rect(fill = "white", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    axis.title        = element_text(face = "bold", size = 14),
    axis.text         = element_text(size = 12),
    axis.line         = element_line(color = "black", linewidth = 0.4),
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(color = "gray85"),
    legend.position   = "none",
    plot.margin       = margin(10,10,10,10)
  ) +
  labs(
    x = "Year",
    y = "Log(Passengers)"
  )
```

---

## 10. Back-Transformation to Original Scale

Given:

$$
\hat{X}_t = \exp(\hat{Y}_t)
$$

Bias-adjusted mean forecast:

$$
E[X_t] = \exp\left(\hat{Y}_t + \frac{\sigma^2}{2}\right)
$$

---

## 11. Train-Test Evaluation

Take last 24 months as the holdout set:

```{r}
h <- 24
train_ts <- window(log_ts, end=c(1958,12))
test_ts  <- window(log_ts, start=c(1959,1))

fit_train <- Arima(train_ts, order=c(0,1,1), seasonal=c(0,1,1))
fcast_train <- forecast(fit_train, h=h)
```

Back-transform:

```{r}
pred <- exp(fcast_train$mean)
actual <- exp(test_ts)
```

Compute error metrics:

```{r}
MAE <- mean(abs(pred - actual))
RMSE <- sqrt(mean((pred - actual)^2))
MAPE <- mean(abs((pred - actual)/actual)) * 100

data.frame(MAE, RMSE, MAPE)
```

---


## 12. Forecast Plots on Original Scale

We now compare forecasts with actual observations (last 24 months):

```{r forecast_vs_actual, message=FALSE, warning=FALSE}

# Reconstruct training and test data on original scale
train_orig <- ts(exp(train_ts), start = start(train_ts), frequency = 12)
test_orig  <- ts(exp(test_ts),  start = start(test_ts),  frequency = 12)
pred_orig  <- ts(pred,          start = start(test_ts),  frequency = 12)

# Plot
ggplot() +
  geom_line(aes(x = time(train_orig), y = train_orig, 
                color = "Training"), linewidth = 1.1) +
  geom_line(aes(x = time(test_orig), y = test_orig,  
                color = "Actual Test"), linewidth = 1.1) +
  geom_line(aes(x = time(pred_orig), y = pred_orig,  
                color = "Forecast"), linewidth = 1.1) +

  scale_color_viridis_d(option = "C") +

  theme_minimal(base_family = "serif") +
  theme(
    panel.background  = element_rect(fill = "white", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    axis.title        = element_text(face = "bold", size = 14),
    axis.text         = element_text(size = 12),
    axis.line         = element_line(color = "black", linewidth = 0.4),
    legend.title      = element_blank(),
    legend.position   = "bottom",
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(color = "gray85"),
    plot.margin       = margin(10,10,10,10)
  ) +
  labs(
    x = "Year",
    y = "Passengers"
  )
```

The model tracks the upward trend and seasonal pattern well.

---

## 13. Mathematical Appendix

### 13.1 SARIMA Model Structure

A general multiplicative SARIMA model is written as:

$$
\Phi_P(B^s)\phi(B) \nabla^d \nabla_s^D X_t
= \Theta_Q(B^s)\theta(B)\varepsilon_t,
$$

where:

- $B$ is the backshift operator
- $\nabla = 1-B$ is the differencing operator
- $\nabla_s = 1-B^s$ is seasonal differencing
- $\phi(B)$ and $\theta(B)$ are nonseasonal AR/MA polynomials
- $\Phi_P(B^s)$ and $\Theta_Q(B^s)$ are seasonal AR/MA polynomials.

For the **Airline Model**:

- $p = 0, d = 1, q = 1$
- $P = 0, D = 1, Q = 1, s = 12$

Thus:

$$
(1-B)(1-B^{12})Y_t = (1 + \theta_1 B)(1 + \Theta_1 B^{12})\varepsilon_t.
$$

---

### 13.2 Forecast Recursion

One-step-ahead forecast:

$$
\hat{Y}_{t+1|t} = 
\sum_{i=1}^p \phi_i Y_{t+1-i}
+ \sum_{j=1}^q \theta_j \hat{\varepsilon}_{t+1-j}
+ \sum_{k=1}^P \Phi_k Y_{t+1-ks}
+ \sum_{\ell=1}^Q \Theta_\ell \hat{\varepsilon}_{t+1-\ell s}.
$$

For multistep forecasts, future residuals $\hat{\varepsilon}_{t+h}$ are set to 0.

---

## 14. Discussion

### Model Performance

- The model captures both **trend** and **seasonality** with only two MA parameters.
- Diagnostics confirm **no residual autocorrelation**, meaning model structure is adequate.
- Forecasts follow the real series closely in the holdout period.

### Strengths

- Simple, interpretable, historically successful model.
- Performs extremely well for multiplicative seasonal series.

### Limitations

- No external regressors included.
- Long-term forecasts may underestimate structural changes.

---

## 15. Recommendations

1. Refit the model periodically as new data arrives.  
2. Consider SARIMAX models if external factors (e.g., economic indicators) are relevant.  
3. For long-term forecasting, evaluate exponential smoothing or structural models.  
4. Use bootstrap intervals for more robust uncertainty quantification.

---

## 16. Conclusion

This report demonstrated a full forecasting workflow, including:

- data transformation  
- decomposition  
- stationarity testing  
- SARIMA identification  
- parameter estimation  
- diagnostics  
- forecasting  
- evaluation  

The canonical Airline Model provides an excellent fit and strong forecasting accuracy.

---

## 17. References

- Box, G.E.P., Jenkins, G.M., Reinsel, G.C., & Ljung, G.M. (2015).  
  *Time Series Analysis: Forecasting and Control.*

- Hyndman, R.J., & Athanasopoulos, G. (2018).  
  *Forecasting: Principles and Practice.*
 

---

